Refactoring the `nls()` function in R

### 2022-05-29 RefactoringNLS/RefactoringNLS.Rmd
add refs to bibliography inc. R-bloggers JN and RB PkgFromRbase22. Edits and improvements to Refactoring

The Jacobian used by `nlsr::nlxb()` is computed from analytic expressions, while that
of `nls()` will use a numerical approximation. In the most recent example above,
the singularity is not due to the numerical approximation.

It should be noted that the selfStart models in the `./src/library/stats/R/zzModels.R` file
provide the Jacobian in the "gradient" attribute of the "one-sided" formula that defines each
model, and these Jacobians may be the analytic forms. The `nls()` function, after computing the
"right hand side" or `rhs` of the residual, then checks to see if the "gradient" attribute is
defined, and, if not, uses `numericDeriv` to compute a Jacobian into that attribute. This code
is within the `nlsModel()` or `nlsModel.pliner()` functions. The use of analytic Jacobians
almost certainly contributes to the performance of `nls()` on selfStart models.

@@

Indeed, the argument only gets used when `nls()` is about to return its result object, and the
element `model` is NULL unless the calling argument `model` is TRUE. (Using the same name could
be confusing.) However, the model frame is used within the function code in the form of an object
`mf`.

@@

Nevertheless, to provide compatible behaviour with `nls()`, we will need to ensure
that equivalent behaviour is guaranteed. Furthermore, we need to test that the 
operation is correct.

@@

Nonlinear estimation algorithms are almost all iterative and need a set of starting
parameters. `nls()` offers a special class of modeling functions called **selfStart** 
models. There are a number of these in base R (see list below) and others in R packages
such as CRAN package `nlraa` (@MiguezNLRAA2021), as well as the now-archived package
`NRAIA`. Unfortunately, the structure of the programming of 
these is such that the methods by which initial parameters are computed is entangled
with the particularities of the `nls()` code. Though there is a `getInitial()` function,
this is not easy to use to simply compute the initial parameter estimates, in part 
because it may call `nls()`. Such circular references are, in our view, dangerous.

In the example below, we show how the `SSlogis` selfStart function can generate a set
of initial parameters for a 3-parameter logistic curve. The form used by `SSlogis`
is $$ y \sim Asym/(1+exp((xmid-tt)/scal)) $$

The example shows how these starting parameters can be transformed to those of 
another form of the model, namely, $$ y \sim b1/(1 + b2*exp(-b3*t)) $$

@@

We note that the function includes analytic expressions for the Jacobian ("gradient").
These could be possibly be useful to R users, especially if documented. Moreover, we
wonder why the programmers have chosen to save so many quantities in "hidden"
variables, i.e., with names preceded by ".". These are then not displayed by the `ls()`
command, making them difficult to access.

@@

Because the Gauss-Newton algorithm is rather unreliable from many starting sets
of parameters, selfStart models are more than an accessory to `nls()` but a part
of the infrastructure. However, creating such functions is a lot of work, and their
documentation (file `./src/library/stats/man/selfStart.Rd`) is quite complicated.
We believe that the focus would better be placed on getting good initial parameters,
possibly with some interactive tools. That is, the emphasis should be on the 
`getInitial()` function, though avoiding the current calls back to `nls()`.

@@

The nls object contains some elements that are awkward to produce by other algorithms.
Moreover, some information that would be useful is not presented obviously.

@@

However, actually looking at the structure of a returned result gives a list of 11
items:
```
 $ resid       : num [1:12] 0.0119 -0.0328 0.092 0.2088 0.3926 ...
  ..- attr(*, "gradient")= num [1:12, 1:3] 0.0271 0.0367 0.0496 0.0666 0.089 ...
  .. ..- attr(*, "dimnames")=List of 2
  .. .. ..$ : NULL
  .. .. ..$ : chr [1:3] "Asym" "xmid" "scal"
 $ jacobian    : num [1:12, 1:3] 0.0271 0.0367 0.0496 0.0666 0.089 ...
  ..- attr(*, "dimnames")=List of 2
  .. ..$ : NULL
  .. ..$ : chr [1:3] "Asym" "xmid" "scal"
 $ feval       : num 31
 $ jeval       : num 23
 $ coefficients: Named num [1:3] 196.19 12.42 3.19
  ..- attr(*, "names")= chr [1:3] "Asym" "xmid" "scal"
 $ ssquares    : num 2.59
 $ lower       : num [1:3] -Inf -Inf -Inf
 $ upper       : num [1:3] Inf Inf Inf
 $ maskidx     : int(0) 
 $ weights     : NULL
 $ formula     :Class 'formula'  language y ~ Asym/(1 + exp((xmid - tt)/scal))
  .. ..- attr(*, ".Environment")=<environment: R_GlobalEnv> 
 - attr(*, "class")= chr "nlsr"
```
This is still a smaller result object than the one `nls()` returns. Moreover, `nlxb`
explicitly returns the sum of squares as well as the residual vector and Jacobian. The 
counts of evaluations are also returned. Working on this project showed several potential
updates to the `nlsr` documentation.

@@

Specifying a model to a solver should, ideally, use the same syntax across
solver tools. Unfortunately, R allows multiple approaches.

@@

However, even within `nls()`, we see annoying inconsistencies. Consider the 
following FOUR different calling sequences for the same problem, though the
second is to illustrate how one intuitive choice will not work. In this failed
attempt, putting the `Asym` parameter in the model causes the `plinear` algorithm
to try to add another term to the model. We believe this is unfortunate, and would
like to see a consistent syntax. At the time of writing (end of August 2021) we do
not have a resolution for this issue.

@@

## using a selfStart model - do not specify the starting parameters
fm1DNase1 <- nls(density ~ SSlogis(log(conc), Asym, xmid, scal), DNase1)
summary(fm1DNase1)
## the coefficients only:
coef(fm1DNase1)
## including their SE, etc:
coef(summary(fm1DNase1))

@@

## Issue: code structure

The `nls()` code is structured in a way that inhibits both maintenance and improvement.
In particular, the iterative setup is such that introduction of Marquardt stabilization
is not easily available.

To obtain performance, a lot of the code is in C with consequent calls and returns that
complicate the code. Over time, R has become much more efficient on modern computers, and
the need to use compiled C and Fortran is less critical. Moreover, the burden for maintenance
could be much reduced by moving code entirely to R.

## Issue: code documentation for maintenance

`setPars()` -- explain weaknesses. Only used by `profile.nls()`

The paucity of documentation is exacerbated by the mixed R/C/Fortran code base.

Following is an email to John Nash from Doug Bates. This is NOT a criticism
of Prof. Bates work, but a reflection on how difficult it is to develop code in this
subject area and to keep it maintainable. We have experienced similar loss of understanding
for some of our own codes.

@@

If the R sources were in a git repository I might try to use `git blame` 
to find out when and by whom that was written but they are in an SVN repository, 
I think, and I haven't used it for a long, long time.

@@

# Goals of our effort

Here are some of the goals we hope to accomplish.

@@

- to provide a packaged version of `nls()` (call it `nlsalt`) coded entirely in R 
  that matches the version in base R or what is packaged in `nlspkg` as described 
  in the "PkgFromRbase" document.

>>>>

The Jacobian used by `nlsr::nlxb()` is computed from analytic expressions, while that
of `nls()` (and, by extension, `minpack.lm::nlsLM`) will use a numerical approximation. 
This can lead to different behaviour:

- A pasture regrowth problem (@Huet2004, page 1, based on @Ratkowsky1983) has a poorly
  conditioned Jacobian and `nls()` fails with "singular gradient". 
  Worse, numerical approximation to the Jacobian give the error
  "singular gradient matrix at initial parameter estimates" for `minpack.lm::nlsLM`
  so that the Marquardt stabilization is unable to take effect, while the analytic
  derivatives of `nlsr::nlxb` give a solution.

- Karl Schilling (private communication) provided an example where a model specified
  with the formula `y ~ a * (x ˆ b)` causes `nlsr::nlxb` to fail because the partial
  derivative w.r.t. `b` is `a * (x^b * log(x))`. If there is data for which `x = 0`,
  this is undefined. In such cases, we observed that `nls()` and `minpack.lm::nlsLM`
  found a solution, though it can be debated whether such lucky accidents can be
  taken as an advantage.

Note that the selfStart models in the `./src/library/stats/R/zzModels.R` file
provide the Jacobian in the "gradient" attribute of the "one-sided" formula that defines each
model, and these Jacobians may be the analytic forms. The `nls()` function, after computing the
"right hand side" or `rhs` of the residual, then checks to see if the "gradient" attribute is
defined, and, if not, uses `numericDeriv` to compute a Jacobian into that attribute. This code
is within the `nlsModel()` or `nlsModel.pliner()` functions. The use of analytic Jacobians
almost certainly contributes to the performance of `nls()` on selfStart models.

@@

Indeed, the argument only gets used when `nls()` is about to return its result object, and the
element `model` is NULL unless the calling argument `model` is TRUE. (Using the same name could
be confusing.) However, the model frame is used within the function code in the form of the object
`mf`.

@@

Nevertheless, to provide compatible behaviour with `nls()`, competing programs 
need to ensure equivalent behaviour, but users should test that the 
operation is correct.

@@

Nonlinear estimation algorithms are almost all iterative and need a set of starting
parameters. `nls()` offers a special class of modeling functions called **selfStart** 
models. There are a number of these in base R (see list below) and others in R packages
such as CRAN package `nlraa` (@MiguezNLRAA2021), as well as the now-archived package
`NRAIA`. Unfortunately, the structure of **selfStart** codes
is such that the methods by which initial parameters are computed is entangled
with the particulars of the `nls()` code. Though there is a `getInitial()` function,
this is not easy to use to simply compute the initial parameter estimates, in part 
because it may call `nls()`. Such circular references are, in our view, dangerous.

In the example below, we show how the `SSlogis` selfStart function can generate a set
of initial parameters for a 3-parameter logistic curve. The form used by `SSlogis`
is $$ y \sim Asym/(1+exp((xmid-tt)/scal)) $$, but we show how the starting parameters
for this model can be transformed to those of 
another form of the model, namely, $$ y \sim b1/(1 + b2*exp(-b3*t)) $$.

@@

This R function includes analytic expressions for the Jacobian ("gradient").
These could be be useful to R users, especially if documented. Moreover, we
wonder why the programmers have chosen to save so many quantities in "hidden"
variables, i.e., with names preceded by ".". These are then not displayed by the `ls()`
command, making them difficult to access.

@@

Because the Gauss-Newton algorithm is unreliable from many starting sets
of parameters, selfStart models are more than an accessory to `nls()` but a part
of the infrastructure. However, creating such functions is a lot of work, and their
documentation (file `./src/library/stats/man/selfStart.Rd`) is quite complicated.
We believe that the focus would better be placed on getting good initial parameters,
possibly with some interactive tools. That is, the emphasis should be on the 
`getInitial()` function, though avoiding the current calls back to `nls()`.

@@

The nls object contains some elements that are awkward to produce by other algorithms,
but some information that would be useful is not presented obviously.

@@

However, actually looking at the structure of a returned result gives a list of 11
items:
```
 $ resid       : num [1:12] 0.0119 -0.0328 0.092 0.2088 0.3926 ...
  ..- attr(*, "gradient")= num [1:12, 1:3] 0.0271 0.0367 0.0496 0.0666 0.089 ...
  .. ..- attr(*, "dimnames")=List of 2
  .. .. ..$ : NULL
  .. .. ..$ : chr [1:3] "Asym" "xmid" "scal"
 $ jacobian    : num [1:12, 1:3] 0.0271 0.0367 0.0496 0.0666 0.089 ...
  ..- attr(*, "dimnames")=List of 2
  .. ..$ : NULL
  .. ..$ : chr [1:3] "Asym" "xmid" "scal"
 $ feval       : num 31
 $ jeval       : num 23
 $ coefficients: Named num [1:3] 196.19 12.42 3.19
  ..- attr(*, "names")= chr [1:3] "Asym" "xmid" "scal"
 $ ssquares    : num 2.59
 $ lower       : num [1:3] -Inf -Inf -Inf
 $ upper       : num [1:3] Inf Inf Inf
 $ maskidx     : int(0) 
 $ weights     : NULL
 $ formula     :Class 'formula'  language y ~ Asym/(1 + exp((xmid - tt)/scal))
  .. ..- attr(*, ".Environment")=<environment: R_GlobalEnv> 
 - attr(*, "class")= chr "nlsr"
```
This is still a smaller result object than the one `nls()` returns. Moreover, `nlxb`
explicitly returns the sum of squares as well as the residual vector and Jacobian. The 
counts of evaluations are also returned. Working on this project showed several potential
updates to the `nlsr` documentation. Note that the singular values of the Jacobian are
actually computed in the `print` and `summary` methods for the result.

@@

Specifying a model to a solver should, ideally, use the same syntax across
solver tools. Unfortunately, R allows multiple approaches within different
modelling tools, and within `nls()` itself.

@@

Within `nls()`, consider the 
following FOUR different calling sequences for the same problem, though the
fm2a -- an intuitive choice -- will not work. In this failed
attempt, putting the `Asym` parameter in the model causes the `plinear` algorithm
to try to add another term to the model. We believe this is unfortunate, and would
like to see a consistent syntax. At the time of writing we do
not foresee a resolution for this issue.

@@

## using a selfStart model - do not specify the starting parameters
fm1DNase1 <- nls(density ~ SSlogis(log(conc), Asym, xmid, scal), DNase1)
summary(fm1DNase1)
## the coefficients only:
# coef(fm1DNase1)
## including their SE, etc:
# coef(summary(fm1DNase1))

@@

## Issue: code structure and documentation for maintenance

The `nls()` code is structured in a way that inhibits both maintenance and improvement.
In particular, the iterative setup is such that introduction of Marquardt stabilization
is not easily available. To obtain performance a lot of the code was in C
with consequent calls and returns that
complicate the code. Over time, R has become much more efficient on modern computers, and
the need to use compiled C and Fortran is less critical. Moreover, the burden for maintenance
could be reduced by moving code entirely to R.

While R and its packages are generally very well documented for usage, that documentation
rarely extends as far as it might in directions helpful for code maintenance. 
The paucity of such documentation is exacerbated by the mixed R/C/Fortran code base,
where seemingly trivial differences in things like representation of numbers or
base index for arrays lead to traps for unwary programmers.

For `nls()` the concern is demonstrated by a short email correspondence.
Following is an email to John Nash from Doug Bates. This is NOT a criticism
of Prof. Bates work, but a reflection on how difficult it is to develop code in this
subject area and to keep it maintainable. We have experienced similar loss of understanding
for some of our own codes.

@@

...

@@

# Goals of our effort to improve `nls()`

@@

- to provide a packaged version of `nls()` (call it `nlsalt`) coded entirely in R 
  that matches the version in base R or what is packaged in `nlspkg` as described 
  in @PkgFromRbase22.
